Supplementary Materials on Bayesian Generalized Additive Models used to Analyse Temporal Trends in Job Ads

### 2025-04-30 analysis/code/010-summary-of-gams.qmd
updates gams supplement

---
title: "Analysis of Temporal Trends using Bayesian Generalized Additive Models"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    code_folding: hide
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
library(gt)          # For potentially nicer tables (optional alternative to kable)
```

@@

Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @R_core]. `brms` utilizes Stan (version `r rstan::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].

@@

# --- Set options for brms (optional, can help speed up) ---
options(mc.cores = parallel::detectCores()) # Use all available CPU cores

@@

# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))
```

@@

Models were fitted using `r brms::nchains(results_brms_valid$mod_brms[[1]])` Markov chains. Each chain consisted of `r results_brms_valid$mod_brms[[1]]$fit@sim$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$fit@sim$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$fit@sim$iter - results_brms_valid$mod_brms[[1]]$fit@sim$warmup) * brms::nchains(results_brms_valid$mod_brms[[1]])` posterior samples per model for inference. No thinning was applied (thin = 1).

@@

Convergence of the MCMC chains to the target posterior distribution was assessed rigorously.

```{r convergence_summary_table, echo=FALSE}
#| tbl-cap: "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0 (e.g., < 1.01). Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0 for reliable results."

@@

```{r fig.height=3 * length(results_brms_valid$mod_brms)}
#| label: fig-trace-plots-mcmc-trace,
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: "Trace plots for key parameters (Intercept and smooth term standard deviation) with divergence markers for each model. Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)     # For map2
library(dplyr)     # For tibble etc.
library(posterior) # For as_draws_array
library(bayesplot) # For mcmc_trace and nuts_params
library(ggplot2)   # For ggtitle and theme adjustments
library(cowplot)   # For plot_grid
#---------------------------------------------------------------------

@@

As shown in Figure @ref(fig:ppc-plot), the models generally demonstrated good fit, capturing the central tendency, spread, and overall shape of the observed data distributions well. The Negative Binomial models effectively captured the skewness of the count data (N. of authors, N. of pages). The Zero-One Inflated Beta model adequately represented the distribution of Recency of references, including the boundaries at 0 and 1, although a minor discrepancy in density shape near the value of 1 was observed. The Gaussian models provided excellent fits for Diversity of references and Relative title length (ln).

@@

Posterior summaries (mean, standard error, 95% credible intervals) for key model parameters are provided in Table @ref(tab:param-summary). This includes intercepts, standard deviations of smooth terms (`sds`), and relevant distributional parameters (`shape`, `sigma`, `phi`, `zoi`, `coi`).

```{r}
#| label: tab:param-summary
#| echo: false
#| message: false
#| warning: false
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)
library(dplyr)
library(tibble) # For rownames_to_column
library(knitr)
library(kableExtra)
#---------------------------------------------------------------------

@@

@@

    # Extract the first plot, add title, label, and theme
    plot_object_list[[1]] +
      ggtitle(var_name) +
      ylab(y_axis_label) +
      theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
  }
)

>>>>

---
title: "Supplementary Materials on Bayesian Generalized Additive Models used to Analyse Temporal Trends in Job Ads"
author: "Ben Marwick"
date: "`r Sys.Date()`"
bibliography: ../analysis/paper/references.bib
csl:          ../analysis/paper/templates/journal-of-archaeological-science.csl # Insert path for the bib-style
output:
  html_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
```

@@

Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @rcoreteamLanguageEnvironmentStatistical2024]. `brms` utilizes Stan (version `r rstan::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].

@@

# --- Set options for brms 
options(mc.cores = parallel::detectCores()) # Use all available CPU cores

@@

# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))

pretty_num <- function(x) format(x, big.mark = ",", scientific = FALSE)

n_chains <- pretty_num(brms::nchains(results_brms_valid$mod_brms[[1]]))
n_total_iter <- pretty_num(results_brms_valid$mod_brms[[1]]$fit@sim$iter)
n_warmup <- pretty_num(results_brms_valid$mod_brms[[1]]$fit@sim$warmup)
n_pos_samples <- 
  pretty_num(
  (results_brms_valid$mod_brms[[1]]$fit@sim$iter - results_brms_valid$mod_brms[[1]]$fit@sim$warmup) * brms::nchains(results_brms_valid$mod_brms[[1]])
  )

@@

Models were fitted using `r n_chains` Markov chains. Each chain consisted of `r n_total_iter` total iterations, with the first `r n_warmup` iterations discarded as warmup, leaving `r n_pos_samples` posterior samples per model for inference. No thinning was applied (thin = 1).

@@

Convergence of the MCMC chains to the target posterior distribution was assessed rigorously (@tbl-convergence-summary-table).

```{r , echo=FALSE}
#| label: tbl-convergence-summary-table
#| tbl-cap: "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0 (e.g., < 1.01). Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0 for reliable results."

@@

```{r fig.height=3 * length(results_brms_valid$mod_brms)}
#| label: fig-trace-plots-mcmc-trace
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: "Trace plots for key parameters (Intercept and smooth term standard deviation) with divergence markers for each model. Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)     # For map2
library(dplyr)     # For tibble etc.
library(posterior) # For as_draws_array
library(bayesplot) # For mcmc_trace and nuts_params
library(ggplot2)   # For ggtitle and theme adjustments
library(cowplot)   # For plot_grid
#---------------------------------------------------------------------

@@

As shown in @fig-ppc-plot, the models generally demonstrated good fit, capturing the central tendency, spread, and overall shape of the observed data distributions well. The Negative Binomial models effectively captured the skewness of the count data (N. of authors, N. of pages). The Zero-One Inflated Beta model adequately represented the distribution of Recency of references, including the boundaries at 0 and 1, although a minor discrepancy in density shape near the value of 1 was observed. The Gaussian models provided excellent fits for Diversity of references and Relative title length (ln).

@@

Posterior summaries (mean, standard error, 95% credible intervals) for key model parameters are provided in @tbl-param-summary. This includes intercepts, standard deviations of smooth terms (`sds`), and relevant distributional parameters (`shape`, `sigma`, `phi`, `zoi`, `coi`).

```{r}
#| label: tbl-param-summary
#| echo: false
#| message: false
#| warning: false
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)
library(dplyr)
library(tibble) # For rownames_to_column
library(knitr)
library(kableExtra)
#---------------------------------------------------------------------

@@

```{r}
#| results: "hide"

@@

    # Extract the first plot, add title, label, and theme
    plot_object_list[[1]] +
      ggtitle(var_name) +
      ylab(y_axis_label) +
      theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
  }
)
```
